Simulation of angular resolution of a new electromagnetic sampling calorimeter

### 2023-03-07 paper.tex
updates

In general, the sampling calorimeter enables cost-effective fabrication of large-scale detectors and design of flexible geometric configuration. This detector has been developed in various ways by selecting a material of absorber among many candidates such as brass~\cite{CMS:mat}, lead~\cite{CDF:mat}, tungsten~\cite{DELPHI:mat}, uranium~\cite{UGAS:mat}, and also of active medium such as scintillators~\cite{CMS:mat,CDF:mat}, gas~\cite{UGAS:mat}, solid-state detector~\cite{DELPHI:mat}, and liquid gas~\cite{LiqAR:mat}. Also, there are many topologies to pile up the materials and read out their signals, such as sandwich~\cite{KOTO:MB}, shashlik~\cite{shashlik:con}, spaghetti~\cite{KLOE:con}, and accordian~\cite{LiqAR:mat}. Sampling calorimeters have been used to measure the direction of arriving photons in a space-based detector to identify astrophysical sources for gamma-ray astrophysics. The PAMELA experiment employs a strip multilayer electromagnetic calorimeter of 22 ($x$,$y$) layers of silicon strip detectors interleaved with 22 tungsten plates, which identifies the particle types and measures the deposited energy, and helps to determine the detection of the particles~\cite{PAMELA}. The Fermi Large Area Telescope (LAT) has a tracker consisting of 18 $x$-$y$ pairs of silicon strip detectors with 16 interleaved tungsten foils. Tracking with the silicon detectors on the electron-positron pairs produced at the foils provides an angular resolution of 0.6 degrees for 1-GeV photons and better than 0.16 degrees for photons with energies higher than 10~GeV~\cite{FERMI:LAT}. The LAT has two additional components: a plastic scintillator for the charged-particle detection and a homogeneous CsI crystals for the energy measurement. This triple detector configuration is also used in different missions. An energy spectrum measurement of cosmic-ray electrons (CALET)~\cite{CALET} utilizes scintillating fibers and PWO crystals instead of the silicon strips and the CsI crystals, respectively. A dark matter particle explorer mission (DMPE)~\cite{DMPE} selects BGO crystals for the energy measurement. Theses detectors provide the angular resolution of 1-2 degrees for 1-GeV photons.

@@

The toy sampling calorimeter is designed as a block consisting of alternating layers of an 1-mm-thick lead absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented into 15-mm-wide strips, which are alternately oriented in the vertical and horizontal directions, as shown in Figure~\ref{fig:det_conf}. It has a cross-section size of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm length and 20 radiation lengths (20$X_{0}$) to contain the EM shower of photons for energies in the range of 0.1 to 2~GeV.

@@

Training data for machine learning are prepared using the Geant4 simulation such that the input dataset is representative of the detector response of the sampling calorimeter. In order to minimize the bias in the training process related to the dataset, the incident angles are uniformly generated at the detector surface in the range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes the azimuthal angle. The number of training samples is $10^{5}$ considering the limited computing resources. The angle reconstruction is studied using photons generated at given incident $\theta$ angle and fixed incident energy and incident position. It turns out to be unchanged with the energy deposit threshold above the noise level of 0.5 MeV. Therefore, we do not consider random noises in this study.

The angular reconstruction can be largely affected by the deviation of the incident position, coming from the spatial resolution of the incident position reconstruction, which can be overcome with the repetitive training for different incident positions. In this study, such effects are not considered when evaluating the angular resolution. The position resolution is expected to be level of 3~mm and to degrade the angular resolution as much as 40\%.

@@

The incident angle of the photons is reconstructed using XGB~\cite{xgboost:2016}, which maps a feature dataset onto a target variable. Here, the feature dataset and the target varialbe correspond to the energy deposit in each scintillator strip and the incident angle of a photon, respectively. The feature dataset size is the number of strips, hence it varies inversely to the strip width to have the same coverage. In the case of the 15-mm-wide strips, 35~$\times$~105 channels compose the detector. XGB constructs gradient-boosted decision trees using the training dataset, efficiently descending the loss function. The hyperparameters of XGB drive how models are trained. In this paper, five main parameters are studied. \textit{N\_estimators} defines the maximum number of allowed decision trees to be developed, and \textit{Max. depth} defines the complexity of the decision-tree structure. \textit{Subsample} controls the fraction of total event samples for each boosting procedure, \textit{learning rate} determines the step length of gradient boosting, and \textit{gamma} regulates the evaluation of each decision tree.

@@

\subsubsection{Number of layers}
We study the dependency of the angle reconstruction on the number of front layers. Figure~\ref{fig:angle_reco_layer} (a) shows the reconstructed angular distribution for 1-GeV photons by using only the front 32 layers, which correspond to 6.2~$X_{0}$. A fraction of the photons fails to be reconstructed because of insufficient information gathering caused by deeply penetrating photons without shower generation, where the fraction is estimated to be 1.9\%. The failed events are represented as a delta function near 0. Such events are removed by requiring the total energy deposit at the front layers to be higher than 10~MeV, which is 1\% of the incident energy. The angular resolution with the front 32 layers is estimated to be 1.30~$\pm$~0.01$^{\circ}$. The evolution of angular resolution degrades with the reduction of the number of layers used, as shown in figure~\ref{fig:angle_reco_layer} (b).

@@

For the toy sampling calorimeter with 15-mm-wide strips and 32 layers, the angular resolution was obtained as 1.30~$\pm$~0.01$^{\circ}$ in machine learning with $10^{5}$ training samples. Energy dependence of the angular resolution can be expressed as $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}}$. For the angular range of $10^{\circ}<\theta<40^{\circ}$, the $p_{0}$ and $p_{1}$ remain unchanged as $p_{0}$ = 0.238 $\pm$ 0.012$^{\circ}$ and $p_{1}$ = 1.248 $\pm$ 0.002$^{\circ}$, respectively.

>>>>

In general, the sampling calorimeter enables cost-effective fabrication of large-scale detectors and design of flexible geometric configuration. This detector has been developed in various ways by selecting a material of absorber among many candidates such as brass~\cite{CMS:mat}, lead~\cite{CDF:mat}, tungsten~\cite{DELPHI:mat}, uranium~\cite{UGAS:mat}, and also of active medium such as scintillators~\cite{CMS:mat,CDF:mat}, gas~\cite{UGAS:mat}, solid-state detector~\cite{DELPHI:mat}, and liquid gas~\cite{LiqAR:mat}. Also, there are many topologies to pile up the materials and read out their signals, such as sandwich~\cite{KOTO:MB}, shashlik~\cite{shashlik:con}, spaghetti~\cite{KLOE:con}, and accordian~\cite{LiqAR:mat}. Sampling calorimeters have been used to measure the direction of arriving photons in a space-based detector to identify astrophysical sources for gamma-ray astrophysics. The PAMELA experiment employs a strip multilayer electromagnetic calorimeter made of 22 ($x$, $y$) silicon strip detectors interleaved with 22 tungsten plates, which identifies the particle types, measures the deposited energy, and helps to determine the direction of the particles~\cite{PAMELA}. The Fermi Large Area Telescope (LAT) has a tracker consisting of 18 $x$-$y$ pairs of silicon strip detectors with 16 interleaved tungsten foils. Tracking with the silicon detectors on the electron-positron pairs produced at the foils provides an angular resolution of 0.6 degrees for 1-GeV photons and better than 0.16 degrees for photons with energies higher than 10~GeV~\cite{FERMI:LAT}. The LAT has two additional components: a plastic scintillator for the charged-particle detection and a homogeneous CsI crystals for the energy measurement. This triple detector configuration is also used in different missions. An energy spectrum measurement of cosmic-ray electrons (CALET)~\cite{CALET} utilizes scintillating fibers and PWO crystals instead of the silicon strips and the CsI crystals, respectively. A dark matter particle explorer mission (DMPE)~\cite{DMPE} selects BGO crystals for the energy measurement. Theses detectors provide the angular resolution of 1-2 degrees for 1-GeV photons.

@@

The toy sampling calorimeter is designed as a block consisting of alternating layers of a 1-mm-thick lead absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented into 15-mm-wide strips, which are alternately oriented in the vertical and horizontal directions, as shown in Figure~\ref{fig:det_conf}. It has a cross-section size of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm length and 20 radiation lengths (20~$X_{0}$) to contain the EM shower of photons for energies in the range of 0.1 to 2~GeV.

@@

Training data for machine learning are prepared using the Geant4 simulation such that the input dataset is representative of the detector response of the sampling calorimeter. In order to minimize the bias in the training process related to the dataset, the incident angles are uniformly generated at the detector surface in the range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes the azimuthal angle. The number of training samples is $10^{5}$ considering the limited computing resources. The angle reconstruction is studied using photons generated at given incident $\theta$ angle and fixed incident energy and position. It turns out to be unchanged with the energy deposit threshold above the noise level of 0.5 MeV, which is sufficiently higher than the expected noise level. Therefore, we do not consider random noises in this study.

@@

The incident angle of the photons is reconstructed using XGB~\cite{xgboost:2016}, which maps a feature dataset onto a target variable. Here, the feature dataset and the target variable correspond to the energy deposit in each scintillator strip and the incident angle of a photon, respectively. The feature dataset size is the number of strips, hence it varies inversely to the strip width to have the same coverage. In the case of the 15-mm-wide strips, 35~$\times$~105 channels compose the detector. XGB constructs gradient-boosted decision trees using the training dataset, efficiently descending the loss function. The hyperparameters of XGB drive how models are trained. In this paper, five main parameters are studied. \textit{N\_estimators} defines the maximum number of allowed decision trees to be developed, and \textit{Max. depth} defines the complexity of the decision-tree structure. \textit{Subsample} controls the fraction of total event samples for each boosting procedure, \textit{learning rate} determines the step length of gradient boosting, and \textit{gamma} regulates the evaluation of each decision tree.

@@

\subsubsection{Number of layers}
We study the dependency of the angle reconstruction on the number of front layers. Figure~\ref{fig:angle_reco_layer} (a) shows the reconstructed angular distribution for 1-GeV photons by using only the front 32 layers, which correspond to 6.2~$X_{0}$. A fraction of the photons fails to be reconstructed because of insufficient information gathering caused by deeply penetrating photons without shower generation, where the fraction is estimated to be 1.9\%. The failed events are represented as a delta function near 0. Such events are removed by requiring the total energy deposit at the front layers to be higher than 10~MeV, which is 1\% of the incident energy. The angular resolution with the front 32 layers is estimated to be 1.30~$\pm$~0.01$^{\circ}$. The evolution of angular resolution degrades with the reduction of the number of layers used, as shown in Figure~\ref{fig:angle_reco_layer} (b).

@@

For the toy sampling calorimeter with 15-mm-wide strips and 32 layers, the angular resolution was obtained as 1.30~$\pm$~0.01$^{\circ}$ in machine learning with $10^{5}$ training samples. The present result can be changed when a photon is incident at a different position from the origin of the training process. Since the position resolution at the current setup is expected to be 2.1~mm, the angular resolution can be degraded by as much as 25\%. Energy dependence of the angular resolution can be expressed as $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}}$. For the angular range of $10^{\circ}<\theta<40^{\circ}$, the $p_{0}$ and $p_{1}$ remain unchanged as $p_{0}$ = 0.238 $\pm$ 0.012$^{\circ}$ and $p_{1}$ = 1.248 $\pm$ 0.002$^{\circ}$, respectively.
